---
title: "Telling a story from data"
author: "Fabrice Mettler"
date: "2025-05-05"
output: html_document
---

```{r setup, include=FALSE}
library(ggplot2)
library(readr)
library(lubridate)
library(dplyr)
library(tidyverse)

airquality <- datasets::airquality # reading in the data
```
## About the dataset
The data set "airquality" contains measurements of the air quality in New York
from May 1, 1973 to September 30, 1973. The data frame contains 153
observations on 6 numeric variables including the mean ozone (ppb) from 1300 
to 1500 hours at Roosevelt Island, solar radiation in Langleys in the frequency
band 4000-7700 Angstroms from 0800 to 1200 hours at Central Park, average wind
speed (miles/h) at 0700 and 1000 hours at LaGuardia Airport, the maximal
temperature (°F) at LaGuardia Airport, the month of the year and the 
day of the month.

I removed NAs and introduced a new column called "Dates".

```{r}
colSums(is.na(airquality)) # checking if there are NAs in the data set
airquality_clean <- airquality |> # removing NAs
  drop_na()

airquality_clean <- airquality_clean |> #  create a new column with the dates
  mutate(
    Month = as.numeric(Month),
    Dates = make_date(year = 1973, month = Month, day = Day))

# airquality_clean$Month <- as.factor(airquality_clean$Month)

knitr::kable(head(airquality_clean)) # presenting an overview of the data set
```

### Question to answer:
What is the temporal pattern of ozone over the analysed time (May - September) 
and how does it correlate with temperature?

### Analysis
To get an overview of the data, I present the range and mean of
measured ozone and the temperature.

```{r}
summary_airquality <- airquality_clean |>
  summarise(
    min_ozone = min(Ozone),
    max_ozone = max(Ozone),
    mean_ozone = mean(Ozone),
    min_temp = min(Temp),
    max_temp = max(Temp),
    mean_temp = mean(Temp)
  )

knitr::kable(summary_airquality, caption = "Some Summaries of the data set")

```

The first plot presents the measured ozone over time:

```{r}
ggplot(airquality_clean, aes(x = Dates, y = Ozone, colour = Temp)) +
  geom_point() +
  geom_line() +  # optional: connects the points in order
  labs(title = "Temporal pattern of ozone", 
       x = "Date",
       y = "Ozone (ppb)",
       colour = "Temperature (°F)") +
  scale_color_gradient(low = "blue", high = "orange") +
  theme_classic()

```


```{r}
ggplot(
  data = airquality_clean,
  aes(x = Temp, y = Ozone, colour = factor(Month))) +
  geom_point(alpha = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  labs(title = "Plot", 
       x = "Temperature (°F)",
       y = "Ozone (ppb)",
       color = "Month") +
  scico::scale_color_scico_d(palette = "romaO")
```

```{r}
ozone_residuals <- lm(Ozone ~ Temp, data = airquality_clean)$residuals


ggplot(
  data = airquality_clean,
  aes(x = Dates, y = Ozone, colour = Temp)) +
  geom_point(alpha = 1) +
  labs(x = "Temperature (F)",
       y = "Ozone (ppb)")
```


