---
title: "Stepwise Forward Regression"
author: "Fabrice Mettler"
date: "2025-05-26"
output: html_document
---

## General informations

```{r setup,  include=FALSE}
library(readr)
library(purrr)
library(tidyverse)
library(ggplot2)
library(scales)
data_stepwise <- read_csv("../data/df_for_stepwise_regression.csv") # loading the data
```
In this report exercise I will perform a step wise forward regression for the task of modelling the GPP (GPP_NT_VUT_REF) as a function of predictors available in the data set. The data set contains air temperature (TA_F), incoming short-wave radiation (SW_IN_F), incoming long-wave radiation (LW_IN_F), vapor pressure deficit (VPD_F), atmospheric pressure (PA), precipitation (P), wind speed (WS), incoming photosynthetic photon flux density (PPFD_IN), CO2 (CO2_F_MDS) and friction velocity (USTAR).

## Evaluation of the bivariate models
For that, I start with bivariate models to evaluate which single predictor fits best.

```{r }
predictors <- data_stepwise |> # defining the predictors
  select(- siteid, -TIMESTAMP, - GPP_NT_VUT_REF) |> 
  names()

bivariate_regressions <- predictors |> # creating a linear regression for each predictor
  set_names() |>
  map(~ lm(as.formula(paste("GPP_NT_VUT_REF", "~", .x)), data = data_stepwise))

bivariate_regressions |>
  map_dbl(~ summary(.x)$r.squared) |>
  enframe(name = "predictor", value = "r_squared") |>
  mutate(r_squared = round(r_squared, 3)) |>
  ggplot2::ggplot(aes(x = reorder(predictor, r_squared), y = r_squared)) +
  geom_col(fill = "blue") +
  geom_text(aes(label = r_squared), hjust = -0.1, size = 3) +
  coord_flip() +
  labs(
    title = "R² of the bivariate models",
    x = "R²", 
    y = "Predictors") +
  theme_minimal() +
  ylim(0, 1)

```

The highest correlation is achieved with the predictor PPFD_IN (R² = 0.452) which is expected as plants use energy of photons and when there are more more photons available, plants can produce more biomass. We can also see that predictors which were gap filled with MDS (end with _MDS) have a very similar R² to the same predictor without gap filling, which is also very intuitive.

## Stepwise forward regression

```{r }
source("../R/stepwise_forward.R")
model_1 <- stepwise_forward(data = data_stepwise, response = "GPP_NT_VUT_REF", predictors = predictors)

lm_result <- tail(model_1$final_model, 1)[[1]]
summary(lm_result)
```


```{r }
rsq_vals <- map_dbl(model_1$final_model, ~ summary(.x)$r.squared)

rsq_plot <- tibble(
  Step = seq_along(rsq_vals),
  R_squared = rsq_vals
) |>
  ggplot(aes(x = Step, y = R_squared)) +
  geom_line(color = "blue") +
  geom_point(color = "black", size = 2) +
  labs(title = "R² per step",
       x = "Step",
       y = "R²") +
  theme_minimal()

rsq_plot
```


```{r }
aic_vals <- map_dbl(model_1$final_model, AIC)

aic_plot <- tibble(
  Step = seq_along(aic_vals),
  AIC = aic_vals
) |>
  ggplot(aes(x = Step, y = aic_vals)) +
  geom_line(color = "blue") +
  geom_point(color = "black", size = 2) +
  labs(title = "AIC per step",
       x = "Step",
       y = "AIC") +
  theme_minimal() +
  scale_y_continuous(labels = label_comma())

aic_plot
```
### Discussion of the results
In the figures above we see the improvement of the R² over the single steps of the step-wise forward regression. R² improves with a saturation curve, which means that the first 4 predictors contribute much to the model improvement while while the rest of the predictors only lead to slight improvement.
In the second plot, the AIC is presented over the single steps of the step-wise regression. Similar to the R², the AIC improves a lot until the first 4 steps when the improvement continues on a low level.
The 


